---
title: "What is best associated with how successful a team has performed in the UEFA Champions League?"
author: "Cole Walker, Madison Griffin"
format: pdf
editor: visual
---

```{r message = F, echo = FALSE, warnings = F}
library(tidyverse)
library(tidymodels)
library(readxl)
library(MASS)
library(leaps)
library(caret)
library(glmnet)
library(Stat2Data)
#library(statnnet)
library(lme4)
library(UpSetR)
library(nlme)
library(sjstats)
set.seed(8)
soccer <- read_excel("AllTimeRankingByClub.xlsx")
```

## Introduction and Data
The UEFA Champions League (UCL) is one of the most entertaining sports 
tournaments in the world. The tournament consists of 32 clubs separated into 
eight groups, and clubs around Europe’s top-division leagues qualify based upon 
where they finish in the final standings from their previous domestic league 
season. Each league is allocated a certain number of qualifying spots based on 
how well their clubs have performed in European continental competitions over 
the previous five years (Snowman Sports Media, 2020). Because of this, clubs 
from Europe’s most talented leagues in England, Germany, Spain, Italy, and 
France typically have 3-4 representatives compete per year, whereas clubs from
smaller, less-talented leagues in countries like Austria, Belgium, and Serbia 
typically only have 1-2 representatives.
 
Winning a UCL title remains one of the most impressive feats in all of sports 
given that each club in the tournament is either a domestic league champion or 
at least one of the other top-four finishers, so the competition is especially 
stiff. In our analysis, we wanted to identify potential statistical predictors 
and how they are associated with a club’s historical success in the UCL. We 
asked the question, what relationships exist between each of these predictors 
and the response variable that we chose to best represent success in the 
tournament? We found a dataset created by Bashar Naji titled 
“AllTimeRankingsByClub” which compiled official data from UEFA’s website and 
highlighted tournament statistics from every club that has competed in the UCL
from its 1955 founding up until the group stage of the 2021/22 season
(Naji, 2022). The dataset incorporated variables including the name of every
club that has competed in the UCL up to the 2021/2022 season, the country 
they belong to, the number of tournaments they have participated in, the total 
number of matches played, the total number of wins, losses, and draws, and the
total number of goals scored, conceded, and overall goal differential 
throughout their history participating in the tournament. 


#### Data Cleaning
Given that the dataset examines more than 60 years of history and that some 
clubs such as juggernauts Real Madrid and Bayern Munich have far more wins and 
higher goal scoring statistics than clubs with far fewer appearances in the 
tournament, we decided to undergo a process of data cleaning to best match this 
reality. First, we mutated the wins statistic to a win-percentage ratio 
(wins/total matches played) to create a more generalized point of comparison 
between the clubs. We then mutated the three goal-related statistics to 
per-match versions with the same intention. In addition, we mutated the club’s 
country league categorical variable into a binary categorical variable that 
differentiated whether the club came from a top-five league (England, Germany, 
Spain, Italy, and France) or not to more clearly discern the competitive 
environments that clubs competed in before qualifying. Lastly, we calculated 
the total number of points per UCL guidelines (3 points for a win and 1 point 
for a draw) that each club has accumulated throughout their UCL history and 
created new “points” and “points per match played” variables to obtain an 
additional measure of success.


```{r message = F, echo = FALSE, warnings = F}
soccer = soccer %>%
  rename(goals_for = `Goals For`,
         goals_against = `Goals Against`,
         goal_diff = `Goal Diff`)

soccer = soccer %>%
  dplyr::select(-Pts)

soccer = soccer %>%
  mutate(winpercentage = Win/Played,
         goalspermatch = goals_for/Played,
         goalsagainstpermatch = goals_against/Played,
         goalmarginpermatch = goal_diff/Played)

soccer = soccer %>%
  mutate(topfiveleague = ifelse(Country == 'ESP' | Country == 'ENG' | Country == 'GER' | Country == 'ITA' | Country == 'FRA', "top five", "not top five"))

soccer = soccer %>%
  mutate(points = (Win * 3) + (Draw * 1))

soccer = soccer %>%
  mutate(pointspermatch = points/Played)

```

#### EDA
We will introduce our final model and the best predictors that it included in 
our Results section. However, this section explains our rationale for choosing 
the initial list of variables that we could include in our model and for 
determining which variable would work best as our response variable and which 
ones would better serve as predictors. Given that only 22 clubs have ever won 
the UCL and the other 508 clubs included in the dataset would have a zero value 
for their titles, we decided to go against the typical practice of using number 
of titles as our response variable. Instead, we determined that the mutated 
“points per match played” variable would provide a more accurate and 
representative determinant of success because it takes into account clubs’ 
overall quality of performance throughout all stages of the tournament. We 
decided to utilize win percentage, the three per-match goal statistics, as well 
as the binary “topfiveleague” variable as the potential predictors. Win 
percentage was a relatively easy inclusion given how essential securing wins 
are for boosting point totals. We included the three per-match goal variables 
because we wanted to examine the relationship between clubs’ offensive prowess, 
defensive brilliance, and overall balance were and their success in the UCL. 
Lastly, we chose the binary “topfiveleague” variable as a predictor because we 
wanted to examine the relationship between the strength of a club’s domestic 
league and their success in the tournament.

Attached below is a visualization showcasing the strong positive relationship 
between a given club’s goals scored per match and its average points accrued 
per match. This visualization represents one example of the linear relationships 
that we found between our chosen predictors and response.

```{r message = F, echo = FALSE, warnings = F}
ggplot(data = soccer, aes(x = goalspermatch, y = pointspermatch)) +
  geom_point() +
  geom_smooth(method = 'lm', se = F, color = 'midnightblue') +
  labs(x = "Goals per Match", y = 'Points per Match', 
       title = 'Strong Positive Relationship between \nGoals per Match and Points per Match') +
  theme_bw()
```

## Methods
Since our outcome variable is numeric and continuous, we knew our model was 
either a linear regression or a linear mixed effects model. We hypothesized that 
the best predictors to determine the success of a club in the UEFA Champions
league were win percentage, goals scored per match, goals scored against per 
match, goal margin per match, and whether a club belonged to a top five country 
league (England, France, Italy, Spain, or Germany).

To select the most effective variables we conducted five different variable 
selection processes: all subset, stepwise (forward, backward, and both), and 
LASSO. 

```{LASSO message = F, echo = FALSE, warnings = F, output = F}
y = soccer$pointspermatch
x = model.matrix(pointspermatch ~ winpercentage + goalspermatch + goalsagainstpermatch +
                 goalmarginpermatch + topfiveleague, data = soccer)
m_lasso_cv = cv.glmnet(x, y, alpha = 1)

best_lambda = m_lasso_cv$lambda.min
best_lambda

m_best = glmnet(x, y, alpha = 1, lambda = best_lambda)
m_best$beta

bestlasso = lm(pointspermatch ~ winpercentage + goalmarginpermatch + topfiveleague,
               data = soccer)
```

```{STEPWISE message = F, echo = FALSE, warnings = F}
m_none = lm(pointspermatch ~ 1, data = soccer)
m_all = lm(pointspermatch ~ winpercentage + goalspermatch + goalsagainstpermatch +
                 goalmarginpermatch + topfiveleague, data = soccer)
```

```{FORWARD message = F, echo = FALSE, warnings = F, output = F}
stepAIC(m_none,
        scope = list(lower = m_none, upper = m_all),
        data = soccer, direction = 'forward')
bestforward = lm(pointspermatch ~ winpercentage + goalmarginpermatch + 
    topfiveleague, data = soccer)
```

```{BACKWARD message = F, echo = FALSE, warnings = F, output = F}
stepAIC(m_all,
        scope = list(lower = m_none, upper = m_all),
        data = soccer, direction = 'backward')
bestbackward = lm(pointspermatch ~ winpercentage + goalspermatch + goalsagainstpermatch +
                    topfiveleague, data = soccer)
```

```{BOTH message = F, echo = FALSE, warnings = F, output = F}
stepAIC(m_none,
        scope = list(lower = m_none, upper = m_all),
        data = soccer, direction = 'both')
bestboth = lm(pointspermatch ~ winpercentage + goalmarginpermatch + topfiveleague, data = soccer)
```

```{ALL message = F, echo = FALSE, warnings = F, output = F}
soccer_allsub = regsubsets(pointspermatch ~ winpercentage + goalspermatch + 
                           goalsagainstpermatch + goalmarginpermatch + topfiveleague,
                           data = soccer,
                           nbest = 1, nvmax = 5)
soccer_allsub 
summary(soccer_allsub)$rsq
summary(soccer_allsub)$which
bestallsubset = lm(pointspermatch ~ winpercentage, data = soccer)
```
The variables selected in forward selection, both directions selection, 
and LASSO were win percentage, goal margin per match, and top five league. The 
variables selected in backward selection were win percentage, goals scored per 
match, goals scored against per match, and top five league. The variable 
selected for all subset selection using Mallo's CP was only win percentage.

### Comparing RMSE after variable selection
To compare each of these models, we compared their RMSE. 

RMSE All Subset = 0.1642128

RMSE Best Backward = 0.1348656

RMSE Best Both, Forward, and Lasso (because they chose the same variables) = 
0.1348667

```{r message = F, echo = FALSE, warnings = F, output = F}
rmse(bestallsubset)
rmse(bestbackward)
rmse(bestboth)
rmse(bestforward)
rmse(bestlasso)
```
Since the model from backwards selection had the lowest RMSE, we decided to use
those variables to assess linear regression assumptions and conditions (win 
percentage, goals scored per match, goals against per match, and top five league).  

We hypothesized that there could be a violation of independence because clubs 
in the same country, especially countries that are in the top 5 league, could 
have access to more money, better facilities, coaches, and training regimens, 
which could violate their independence from each other. Because of this, we 
tested the linear model assumptions and conditions using the variables chosen 
by backward selection in two models: 1) a linear regression, and 2) a linear 
mixed effects model with a random intercept for top five leagues.

#### Checking Assumptions
Model 1: Linear Regression

**Outcome:**

- points per match

**Predictors:**

- win percentage

- goal scored per match

- goals against per match

- top five league
```{r message = F, echo = FALSE, warnings = F}
linear = lm(pointspermatch ~ winpercentage + goalspermatch + goalsagainstpermatch + 
              topfiveleague, data = soccer)
```
For the linear regression, the residual plot (shown below) violates both 
linearity and constant variance. The model starts to underpredict more on the 
right side of the graph, and there are three diagonal patterns across the 
residual plot. The Q-Q plot (shown below) also deviates from the line in the 
bottom left and upper right, thus violating normality.
```{r message = F, echo = FALSE, warnings = F}
linearaug = augment(linear)

ggplot(linearaug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = 'maroon') +
  labs(x = "Fitted (predicted) values", y = 'Residual') +
  ggtitle('Residual Plot Violates Linearity & Constant Variance Assumptions') +
  theme_bw()

ggplot(linearaug, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  theme_bw() +
  labs(x = 'Theoretical quantiles',
       y = 'Sample quantiles',
       title = 'QQ Plot Violates Normality Assumption')
```
Model 2: Linear Mixed Effects Model

**Outcome:**

- points per match

**Predictors:**

- win percentage

- goal scored per match

- goals against per match

- random intercept for top five league
```{r message = F, echo = FALSE, warnings = F}
linearmixed = lmer(pointspermatch ~ 1 + winpercentage + goalspermatch + goalsagainstpermatch +
              (1|topfiveleague), data = soccer)
```
The residual plot (shown below) for the linear mixed effects model also has 
three diagonal patterns. The residual plot is also clumped, and begins to 
underpredict on the right side of the plot. The histogram of residuals 
(shown below) also violates normality, as the bins are relatively large and 
the bars in the middle fall outside of the normal curve.
```{r message = F, echo = FALSE, warnings = F}
plot(linearmixed, xlab = "Fitted (predicted) values", ylab = 'Residual', main = 'Residual Plot Violates Linearity and Constant Variance')
hist((resid(linearmixed) - mean(resid(linearmixed))) / sd(resid(linearmixed)), xlab = "Residual", main = "Histogram of Residuals Violates Normality", freq = FALSE); curve(dnorm, add = TRUE) 

```
Both models violated the assumptions, however, with the potential violation of 
independence because of the variable top five league, we will choose the 
linear mixed model as our final model.

## Results

**FINAL MODEL:** 

$y_{ij}$ = ($\gamma_{00}$ + $\mu_{0j}$) + $\gamma_1 WinPercentage_{ij}$ + 
$\gamma_2 TopFiveLeague_{ij}$ + $\gamma_3 GoalsPerMatch_{ij}$ +
$\gamma_4 GoalsAgainstperMatch_{ij}$ + $\epsilon_{ij}$

where

$y_{ij}$ = points per match

$\gamma_1$: wins percentage

$\gamma_2$: top five league, 1 = top five

$\gamma_3$: goals per match

$\gamma_4$: goals against per match

```{r message = F, echo = FALSE, warnings = F}
finalmodel = lmer(pointspermatch ~ 1 + winpercentage + goalspermatch + goalsagainstpermatch +
              (1|topfiveleague), data = soccer)
summary(finalmodel)
```
The coefficient for win percentage (fixed effect) is 2.383. This means that at a 
given league (top five or not top five), every additional one-unit increase in a 
given club’s win percentage, their predicted points per match is expected to 
increase by 2.383 points per match, while controlling for other variables in our 
model. The coefficient for goals against per match is -0.0984. This means that 
at a given league (top five or not top five), for every additional goal against 
per match, points per match is expected to decrease by 0.0984, while controlling 
for other variables in our model.  

Though the output does not provide p-values, we will refer to t values to 
interpret significance of our fixed effects. Both win percentage (t value = 
19.082) and goals against per match (t value = -14.843) have high t values, 
indicating their significance. Goals scored per match have a small t values
(t value = 6.688), showing it was not as significant as the other predictors.

In summary, win percentage, a club’s defensive ability, a club’s offensive 
ability, and whether a club is from a country in the top five leagues were 
found to be the best predictors of UEFA Champions League success, however win 
percentage and a club’s defensive ability were the most significant.

## Conclusion
In summary, our analysis allowed us to examine the positive and negative 
relationships between potential statistical predictors associated with a given 
club’s historical success in the UCL. Our model found that each of the three 
fixed predictors chosen by our variable selection methods had a linear 
relationship with our chosen response variable that best represented success 
in the tournament. In statistical language, regardless of the club’s domestic 
league, our model predicted that there was a positive relationship between both
a club’s win percentage and their predicted points per match as well as between 
their goals per match and their predicted points per match, while adjusting for
the other variables in the model. Our model also predicted that, regardless of 
the club’s domestic league, there was a negative relationship between a club’s 
goals against per match stat and their predicted points per match, while 
adjusting for the other variables in the model. These findings match real-world 
expectations given that there is a three-point gain for winning UCL matches and 
that maximizing goals scored and minimizing goals allowed attributes to winning 
and thus gaining points. By randomizing the effect of a club belonging to top 
five league or non-top-five league, we were able to address potential violations 
of independence between clubs given that clubs from the same league or tiered 
leagues may have similar access to resources, training equipment, and quality of
coaching personnel.
 
Unfortunately, some clear limitations still exist in our data and two especially
stand out. First, our original dataset did not provide any additional information 
on which specific years each team has competed in the UCL, so we were unable to 
determine whether or not their statistics incorporated data from the pre-1992 
tournament rebranding and subsequent gradual rule/qualification changes, which 
could have affected the rigors of the tournament. Second, as was referenced in 
the introduction, we utilized points per game as our response variable that best 
attributed success in the UCL rather than titles won to better incorporate all 
our observations even though the latter is almost always seen as the best 
determinant of success in any sports tournament.
 
Future studies could be designed to address some of these limitations. First,
more rigorous research could be conducted to incorporate each club’s 
participating seasons in the UCL into the dataset to account for potential rule
changes over time. Second, a future dataset could incorporate a new variable that 
averages how long. Second, perhaps a study only incorporating the clubs who 
have won the UCL could examine the relationship between the same predictors we 
chose and a new response variable called “title success” (number of titles 
divided by number of tournaments participated in). This analysis could enable 
researchers to better understand how factors such as offensive prowess, 
defensive stoutness, and the ability to secure wins each associate with a team’s 
title success.

## Appendix

The plot below shows the number of titles per club, excluding clubs that have
never won a title. As shown here, this would severely cut down our dataset to 
only 22 observations, which is why we decided to not define number of titles 
as "success".
```{r message = F, echo = FALSE, warnings = F}
soccer %>%
  filter(Titles > 0) %>%
  ggplot(aes(x = reorder(Club, (-Titles)), y = Titles)) +
  geom_bar(stat = 'identity') +
  labs(x = 'Club', y = 'Number of Titles', title = 'Number of Titles for each Club',
       subtitle = 'Excluding Clubs That Have Never Won a Title') +
  theme_bw() +
  scale_x_discrete(guide = guide_axis(angle = 45))
```

The plot below shows the number of titles per country, excluding clubs that 
have never won a title. This shows that Spain, England, Germany, Italy, and the
Netherlands have clubs that have won the most UCL titles. Spain, England, 
Germany, Italy, and France are all considered the most talented leagues, 
showing why we created a variable to separate the top five leagues from other
countries.
```{r message = F, echo = FALSE, warnings = F}
soccer %>%
  filter(Titles > 0) %>%
  ggplot(aes(x = reorder(Country, (-Titles)), y = Titles)) +
  geom_bar(stat = 'identity') +
  labs(x = 'Country', y = 'Number of Titles', title = 'Number of Titles per Country',
       subtitle = 'Excluding Clubs That Have Never Won a Title') +
  theme_bw() +
  scale_x_discrete(guide = guide_axis(angle = 45))
```



















