---
title: "What Best Predicts Success of a Club in UEFA Champions League?"
author: "Cole Walker, Madison Griffin"
format: pdf
editor: visual
---

```{r message = F, echo = FALSE, warnings = F}
library(tidyverse)
library(tidymodels)
library(readxl)
library(MASS)
library(leaps)
library(caret)
library(glmnet)
library(Stat2Data)
#library(statnnet)
library(lme4)
library(UpSetR)
library(nlme)
library(sjstats)
set.seed(8)
soccer <- read_excel("AllTimeRankingByClub.xlsx")
```

## Introduction and Data

#### Data Cleaning
```{r message = F, echo = FALSE, warnings = F}
soccer = soccer %>%
  rename(goals_for = `Goals For`,
         goals_against = `Goals Against`,
         goal_diff = `Goal Diff`)

soccer = soccer %>%
  dplyr::select(-Pts)

soccer = soccer %>%
  mutate(winpercentage = Win/Played,
         goalspermatch = goals_for/Played,
         goalsagainstpermatch = goals_against/Played,
         goalmarginpermatch = goal_diff/Played)

soccer = soccer %>%
  mutate(topfiveleague = ifelse(Country == 'ESP' | Country == 'ENG' | Country == 'GER' | Country == 'ITA' | Country == 'FRA', "top five", "not top five"))

soccer = soccer %>%
  mutate(points = (Win * 3) + (Draw * 1))

soccer = soccer %>%
  mutate(pointspermatch = points/Played)

```

#### EDA

Plot 3:
```{r message = F, echo = FALSE, warnings = F}
ggplot(data = soccer, aes(x = goalspermatch, y = pointspermatch)) +
  geom_point() +
  geom_smooth(method = 'lm', se = F, color = 'midnightblue') +
  labs(x = "Goals per Match", y = 'Points per Match', 
       title = 'Strong Positive Relationship between \nGoals per Match and Points per Match') +
  theme_bw()
```

## Methods
Since our outcome variable is numeric and continuous, we knew our model was 
either a linear regression or a linear mixed effects model. We hypothesized that 
the best predictors to determine the success of a club in the UEFA Champions
league were win percentage, goals scored per match, goals scored against per 
match, goal margin per match, and whether a club belonged to a top five country 
league (England, France, Italy, Spain, or Germany).

To select the most effective variables we conducted five different variable 
selection processes: all subset, stepwise (forward, backward, and both), and 
LASSO. 

```{LASSO message = F, echo = FALSE, warnings = F, output = F}
y = soccer$pointspermatch
x = model.matrix(pointspermatch ~ winpercentage + goalspermatch + goalsagainstpermatch +
                 goalmarginpermatch + topfiveleague, data = soccer)
m_lasso_cv = cv.glmnet(x, y, alpha = 1)

best_lambda = m_lasso_cv$lambda.min
best_lambda

m_best = glmnet(x, y, alpha = 1, lambda = best_lambda)
m_best$beta

bestlasso = lm(pointspermatch ~ winpercentage + goalmarginpermatch + topfiveleague,
               data = soccer)
```

```{STEPWISE message = F, echo = FALSE, warnings = F}
m_none = lm(pointspermatch ~ 1, data = soccer)
m_all = lm(pointspermatch ~ winpercentage + goalspermatch + goalsagainstpermatch +
                 goalmarginpermatch + topfiveleague, data = soccer)
```

```{FORWARD message = F, echo = FALSE, warnings = F, output = F}
stepAIC(m_none,
        scope = list(lower = m_none, upper = m_all),
        data = soccer, direction = 'forward')
bestforward = lm(pointspermatch ~ winpercentage + goalmarginpermatch + 
    topfiveleague, data = soccer)
```

```{BACKWARD message = F, echo = FALSE, warnings = F, output = F}
stepAIC(m_all,
        scope = list(lower = m_none, upper = m_all),
        data = soccer, direction = 'backward')
bestbackward = lm(pointspermatch ~ winpercentage + goalspermatch + goalsagainstpermatch +
                    topfiveleague, data = soccer)
```

```{BOTH message = F, echo = FALSE, warnings = F, output = F}
stepAIC(m_none,
        scope = list(lower = m_none, upper = m_all),
        data = soccer, direction = 'both')
bestboth = lm(pointspermatch ~ winpercentage + goalmarginpermatch + topfiveleague, data = soccer)
```

```{ALL message = F, echo = FALSE, warnings = F, output = F}
soccer_allsub = regsubsets(pointspermatch ~ winpercentage + goalspermatch + 
                           goalsagainstpermatch + goalmarginpermatch + topfiveleague,
                           data = soccer,
                           nbest = 1, nvmax = 5)
soccer_allsub 
summary(soccer_allsub)$rsq
summary(soccer_allsub)$which
bestallsubset = lm(pointspermatch ~ winpercentage, data = soccer)
```
The variables selected in forward selection, both directions selection, 
and LASSO were win percentage, goal margin per match, and top five league. The 
variables selected in backward selection were win percentage, goals scored per 
match, goals scored against per match, and top five league. The variable 
selected for all subset selection using Mallo's CP was only win percentage.

### Comparing RMSE after variable selection
To compare each of these models, we compared their RMSE. 

RMSE All Subset = 0.1642128

RMSE Best Backward = 0.1348656

RMSE Best Both, Forward, and Lasso (because they chose the same variables) = 
0.1348667

```{r message = F, echo = FALSE, warnings = F, output = F}
rmse(bestallsubset)
rmse(bestbackward)
rmse(bestboth)
rmse(bestforward)
rmse(bestlasso)
```
Since the model from backwards selection had the lowest RMSE, we decided to use
those variables to assess linear regression assumptions and conditions (win 
percentage, goals scored per match, goals against per match, and top five league).  

We hypothesized that there could be a violation of independence because clubs 
in the same country, especially countries that are in the top 5 league, could 
have access to more money, better facilities, coaches, and training regimens, 
which could violate their independence from each other. Because of this, we 
tested the linear model assumptions and conditions using the variables chosen 
by backward selection in two models: 1) a linear regression, and 2) a linear 
mixed effects model with a random intercept for top five leagues.

#### Checking Assumptions
Model 1: Linear Regression

**Outcome:**

- points per match

**Predictors:**

- win percentage

- goal scored per match

- goals against per match

- top five league
```{r message = F, echo = FALSE, warnings = F}
linear = lm(pointspermatch ~ winpercentage + goalspermatch + goalsagainstpermatch + 
              topfiveleague, data = soccer)
```
For the linear regression, the residual plot (shown below) violates both 
linearity and constant variance. The model starts to underpredict more on the 
right side of the graph, and there are three diagonal patterns across the 
residual plot. The Q-Q plot (shown below) also deviates from the line in the 
bottom left and upper right, thus violating normality.
```{r message = F, echo = FALSE, warnings = F}
linearaug = augment(linear)

ggplot(linearaug, aes(x = .fitted, y = .resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = 'maroon') +
  labs(x = "Fitted (predicted) values", y = 'Residual') +
  ggtitle('Residual Plot Violates Linearity & Constant Variance Assumptions') +
  theme_bw()

ggplot(linearaug, aes(sample = .resid)) +
  stat_qq() +
  stat_qq_line() +
  theme_bw() +
  labs(x = 'Theoretical quantiles',
       y = 'Sample quantiles',
       title = 'QQ Plot Violates Normality Assumption')
```
Model 2: Linear Mixed Effects Model

**Outcome:**

- points per match

**Predictors:**

- win percentage

- goal scored per match

- goals against per match

- random intercept for top five league
```{r message = F, echo = FALSE, warnings = F}
linearmixed = lmer(pointspermatch ~ 1 + winpercentage + goalspermatch + goalsagainstpermatch +
              (1|topfiveleague), data = soccer)
```
The residual plot (shown below) for the linear mixed effects model also has 
three diagonal patterns. The residual plot is also clumped, and begins to 
underpredict on the right side of the plot. The histogram of residuals 
(shown below) also violates normality, as the bins are relatively large and 
the bars in the middle fall outside of the normal curve.
```{r message = F, echo = FALSE, warnings = F}
plot(linearmixed, xlab = "Fitted (predicted) values", ylab = 'Residual', main = 'Residual Plot Violates Linearity and Constant Variance')
hist((resid(linearmixed) - mean(resid(linearmixed))) / sd(resid(linearmixed)), xlab = "Residual", main = "Histogram of Residuals Violates Normality", freq = FALSE); curve(dnorm, add = TRUE) 

```
Both models violated the assumptions, however, with the potential violation of 
independence because of the variable top five league, we will choose the 
linear mixed model as our final model.

## Results

**FINAL MODEL:** 

$y_{ij}$ = ($\gamma_{00}$ + $\mu_{0j}$) + $\gamma_1 WinPercentage_{ij}$ + 
$\gamma_2 TopFiveLeague_{ij}$ + $\gamma_3 GoalsPerMatch_{ij}$ +
$\gamma_4 GoalsAgainstperMatch_{ij}$ + $\epsilon_{ij}$

where

$y_{ij}$ = points per match

$\gamma_1$: wins percentage

$\gamma_2$: top five league, 1 = top five

$\gamma_3$: goals per match

$\gamma_4$: goals against per match

```{r message = F, echo = FALSE, warnings = F}
finalmodel = lmer(pointspermatch ~ 1 + winpercentage + goalspermatch + goalsagainstpermatch +
              (1|topfiveleague), data = soccer)
summary(finalmodel)
```
The coefficient for win percentage (fixed effect) is 2.383. This means that at a 
given league (top five or not top five), every additional one-unit increase in a 
given club’s win percentage, their predicted points per match is expected to 
increase by 2.383 points per match, while controlling for other variables in our 
model. The coefficient for goals against per match is -0.0984. This means that 
at a given league (top five or not top five), for every additional goal against 
per match, points per match is expected to decrease by 0.0984, while controlling 
for other variables in our model.  

Though the output does not provide p-values, we will refer to t values to 
interpret significance of our fixed effects. Both win percentage (t value = 
19.082) and goals against per match (t value = -14.843) have high t values, 
indicating their significance. Goals scored per match have a small t values
(t value = 6.688), showing it was not as significant as the other predictors.

In summary, win percentage, a club’s defensive ability, a club’s offensive 
ability, and whether a club is from a country in the top five leagues were 
found to be the best predictors of UEFA Champions League success, however win 
percentage and a club’s defensive ability were the most significant.

## Conclusion

## Appendix

Plot 1: 
```{r message = F, echo = FALSE, warnings = F}
soccer %>%
  filter(Titles > 0) %>%
  ggplot(aes(x = reorder(Club, (-Titles)), y = Titles)) +
  geom_bar(stat = 'identity') +
  labs(x = 'Club', y = 'Number of Titles', title = 'Number of Titles for each Club',
       subtitle = 'Excluding Clubs That Have Never Won a Title') +
  theme_bw() +
  scale_x_discrete(guide = guide_axis(angle = 45))
```

Plot 2:
```{r message = F, echo = FALSE, warnings = F}
soccer %>%
  filter(Titles > 0) %>%
  ggplot(aes(x = reorder(Country, (-Titles)), y = Titles)) +
  geom_bar(stat = 'identity') +
  labs(x = 'Country', y = 'Number of Titles', title = 'Number of Titles per Country',
       subtitle = 'Excluding Clubs That Have Never Won a Title') +
  theme_bw() +
  scale_x_discrete(guide = guide_axis(angle = 45))
```



















